---
title: "An R package for synthesizing spatiotemporal point patterns for social and life science research - A user guide"

author: |
  | Monsuru Adepeju^[Big Data Centre, Manchester Metropolitan University, Manchester, M156BH, UK, m.adepeju@mmu.ac.uk]
  
date: |
  | `Date:`
  | ``r Sys.Date()``

#output:
  #pdf_vignette
  
output:
  pdf_document

urlcolor: blue

toc:
  true

number_sections: 
  yes
    
linestretch: 
  1.5
  
fontsize: 
  16pt

#header-includes:
 # - \usepackage{leading}
  #- \leading{18}
  
#dev: png
#output:
  #word_document: default
  #always_allow_html: yes
#  pdf_document: default
always_allow_html: yes
#fig_caption: yes
bibliography: references.bib

abstract: Generates artificial spatio-temporal (ST) point patterns 
  through the integration of microsimulation and agent-based models. 
  Allows a user to define the behaviours of a set of 'walkers' (agents, 
  objects, persons, etc,) whose interactions with the spatial (landscape) 
  and the temporal domains produce new point events. The resulting ST 
  patterns from the point cloud can be measured and utilized for spatial 
  and/or temporal model testings and evaluations. With 
  increasingly limited availability of fine-grained spatially and 
  temporally stamped point data, the package provides an alternative 
  source of data for a wide range of fields in social and life sciences.

vignette: >
  %\VignetteIndexEntry{An R package for synthesizing spatiotemporal point patterns for social and life science research - A user guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r functions, include=FALSE}
# A function for captioning and referencing images
fig <- local({
  i <- 0
  ref <- list()
  list(
    cap=function(refName, text) {
      i <<- i + 1
      ref[[refName]] <<- i
      paste("Figure ", i, ": ", text, sep="")
    },
    ref=function(refName) {
      ref[[refName]]
    })
})
```

## Introduction

In many research context, access to fine-grained spatiotemporal (ST) point data have been severely restricted due to privacy concerns. The `R-stppSim` package has been designed to address this challenge by presenting a framework that is capable of mimicking a real-life data through the integration of microsimulation and agent-based techniques. The framework comprises a set of 'walkers' (agents, objects, persons, etc.) with modifiable movement and interaction properties, constructed within specified spatial and temporal domains. As the framework produces new point events, in accordance with a user setting, a defined ST pattern emerges at both the local and global levels.

The package contains two main functions for synthesizing new datasets; (i) `psim_artif` and (ii) `psim_real`. The `psim_artif` synthesizes ST point patterns from scratch. That is, the simulation is entirely based on a user's setting. On the other hand, `psim_real` synthesizes ST point patterns based on a sample (or a sparse version) of a real data. The function first learns (or extracts) the certain ST characteristics of the sample data, and extrapolates to produce full dataset. The potential application fields for `stppSim` include criminology, epidemiology, and wildlife science.

## Simulation Parameters

The simulation parameters are in relation to three elements, namely; the '`Walkers (agents)`', the `landscape` (spatial), and the `time dimension`. The parameters are described as follow: 

### Walkers (agents).

The walkers are defined primarily by the following characteristics:

* ***Origins*** - The walkers emanate from a set of origins distributed randomly across the landscape or origins determined from a sample of real point data. Origins are provided as a set of `xy` coordinates (`coords`). In the analysis of human offenders, as an example, origins are synonymous to offender residences. The spatial pattern of `origins` can be either of two concentration types. First, is the `nucleated` in which all origins concentrate around one focal point, and the second, the `dispersed` in which there is no focal point.
* ***Movement*** - Walkers are configured to move in any direction and also be aware of restrictions on their path. The movements are controlled primarily by an in-built transition matrix (TM) comprising two transitional states, namely; an `exploratory` state (in which a walker is merely exploring the environment) and a `performative` state (in which a walker is performing an action). The stochastic properties of the TM ensure variations in the behavioural patterns amongst the walkers.
* ***Spatial threshold*** - Distance that defines the perception range of a walker at any location. The perception range is generally updated as a walker moves to a new location. Specified when data is simulated from scratch or learnt when data is simulated from a sample real data.
* ***Steps*** - The maximum step taken by a walkers as they move from one location to another. This impacts the spread of point patterns across the space. The `steps` should be carefully defined, especially, when movements of walkers are restricted to narrow paths, such as route network, in which case the `steps` should be smaller than the path widths. 
* ***Proportional ratios*** - Controls the percentage of total events emanating from a small number of origins (i.e., most active origins). 


### Landscape

The followings are key properties of the landscape:

* ***Boundary*** - Defined by a polygon shapefile (`poly`) or by the spatial extent of a sample point data.
* ***Restrictions*** - Provided as a raster layer. A restriction map would normally show areas outside the boundary as having the maximum restriction value of `1` (that is, no movement is allowed). Furthermore, it would show features within the boundary, such as land use or urban structures, that serve as obstructions to movements. The values may range between 0 and 1.
* ***Focal points*** - Synonymous to city/town centres (in urban studies) which usually contain relatively higher concentration of events. Areas around focal points present more opportunities to event occurrences. 

### Time dimension

The following parameters define the time dimension:

* ***global pattern*** - Defines the temporal variation in events over time. Its composed of two aspects: `long-term trend` and `seasonality`. The long-term trend can be `stable`, `increasing` or `decreasing`. The `seasonality` is controlled by the parameter; `first_pDate`, indicating the date of the first seasonal peak of a time series. A `90` day first peak implies a seasonal cycle of `180` days.
* ***time bin*** - Time range to reset all walkers.


## Installation of `stppSim`

From an R console, type:

```{r eval=FALSE, message=FALSE, warning=FALSE}
install.packages("stppSim")
#To install the developmental version of the package, type:
remotes::install_github("MAnalytics/stppSim")
```

Note: `remotes` is an extra package that needed to be installed prior to the running of this code. 

```{r eval=FALSE, message=FALSE, warning=FALSE}
#Now, load the package,
library(stppSim)
```


## Synthesizing `stpp` from scratch

To simulate point patterns from scratch, the argument `n_events` specifies the number of points to simulate. It is recommended to input a vector of values, instead of a single value. For example, `n_events = c(200, 500, 1000, 2000)`. This saves the user the time from having to synthesize new data. ***Note***: the specification of `n_events` input has little or no impacts on the computational time (See the manual for further details). 

### Example using a restricted landscape

Given a boundary shapefile {e.g., data(`camden_boundary`} and land use features with restriction values {e.g., data(`landuse`) - `Leisure` (0.5); `Sports` (0.7); and `Green` (0.9)}, the ST point pattern can be generated as follows: 

```{r eval=FALSE, message=FALSE, warning=FALSE}
#Using datasets that come with the package;
data(camden_boundary)
data(landuse)

#specifyings say 3 data sizes
pt_sizes = c(200, 1000, 2000)

artif_stpp <- psim_artif(n_events=pt_sizes, start_date = "2021-01-01",
  poly=camden_boundary, n_origin=50, resistance_feat = landuse,
  field = "rValue1",
  n_foci=5, foci_separation = 10, conc_type = "dispersed",
  p_ratio = 20, s_threshold = 50, step_length = 20,
  trend = "stable", first_pDate=NULL,
  slope = NULL,show.plot=FALSE, show.data=FALSE)
```

To preview the output:

```{r eval=FALSE}
head(artif_stpp)
```

Users should explore the impacts of different arguments, including the `n_origin`, `foci_separation`, `trend` {`stable`,`increasing` or `decreasing`},  and `conc_type` {`nucleated` or `dispersed`}.  

Exploring the results using visualization:

* **Figure 1abc - spatial patterns of 200, 1000. 2000 data points**

The corresponding temporal pattern of the data can be visualized as follows:

* **Figure 2 - temporal patterns of 200, 1000. 2000 data points**


## Synthesizing `stpp` from a sample real dataset

Here, we are going to extract 30% random sample the `Theft` crime of Camden, then utilize the sample to synthesize a `full` data size.

```{r eval=FALSE, message=FALSE, warning=FALSE}
data(camden_crimes)

#get the 'theft' crime
theft <- camden_crimes %>%
  filter(type == "Theft")

#specify the proportion to extract
sample_size <- 0.3 #i.e. 30% of real data

set.seed(1000)
dat_sample <- theft[sample(1:nrow(theft),
  round((sample_size * nrow(theft)), digits=0),
  replace=FALSE),1:3]

#As a user would not normally know the actual size of 
#the real full data, therefore, we would assume `n_events`
#to be `2000`. In practice, a user should infer 
#the input size from any available (similar) data 
#from the study area.

#Now, simulate
sim_fullData <- psim_real(n_events=2000, ppt=dat_sample,
  start_date = NULL, poly = NULL, s_threshold = NULL,
  step_length = 20, n_origin=50, resistance_feat, field=NA,
  p_ratio=20, crsys = "EPSG:27700")

#plot(dat_sample$x, dat_sample$y) #preview

```

Preview the resulting data using visualization approach (above).

### Comparing simulated data and full real data

We will utilize a `visual approach` and a `statistical approach` to compare the spatial and temporal patterns of the simulated data and the `100%` real data size. The visual approach involves the comparison of `heat maps` derived from the aggregation of the datasets to a square grid system (250m x 250m). The statistical approach involves computing the `pearson coefficient` to compares the two datasets.

* Visual approach

** Spatial
Figure 1a (heat map of simulated data) 
Figure 1b (heat map of full real data)

** Temporal
Figure 2a (Time series plot of simulated data) 
Figure 2b (Time series plot of full real data)

* Statistical approach

Flattening the 2D square grids (above) and then apply Pearson correlation:

```{r eval=FALSE}
#corr

cor(sim, real)

```


## Discussion

This vignette has demonstrated the basic utility of `stppSim` package for generating fine-grained spatiotemporal point patterns, (1) from scratch and (2) from a sample of real data sets. The parameters of the three simulation elements: the 'Walkers (agents)', the landscape (spatial), and the time dimension should be modified in accordance with the research questions at hand. The package has a wide potential areas of applications. Examples of objects with similar objectives as walkers (as defined in this package) include, human offenders, foraging animals and disease carriers, in victimization, Wildlife, and epidemiological studies, respectively. We will continue to update the package for wider application.   

We encourage users to report any bugs encountered while using the package so that they can be fixed immediately. Welcome contributions to this package which will be acknowledged accordingly. 
